Dynamic Witnesses for Static Type Errors

### 2015-11-20 paper/pldi16-submission/evaluation.tex
more WRW comments
replace pdf figures with pngs

\begin{figure}[t]
\centering
\begin{minipage}{\linewidth}
\includegraphics[width=\linewidth]{coverage.pdf}
\end{minipage}
\begin{minipage}{\linewidth}
\includegraphics[width=\linewidth]{distrib.pdf}
\end{minipage}
% \vspace{-8ex}
\caption{Results of our coverage testing and the distribution of test
  outcomes. Our random search successfully finds witnesses for 83\% of
  the programs in under one second, improving slightly to 87\% in under
  10 seconds. In both datasets we detect actual type errors about 82\%
  of the time, unbound variables or constructors 3-4\% of the time, and
  diverging loops 2-3\% of the time. For the remaining 11-12\% of the
  programs we are unable to provide any useful feedback.  }
\label{fig:results-witness}
\end{figure}

@@

\paragraph{Jump Compression}
%
A jump compressed trace is one whose edges are limited to forward
or backward jumps. (The sizes of these traces is equal to the jump
metric below.)
%
In our experience, jump compression abstracts many details of the
computation that are often uninteresting or irrelevant to the
explanation, or at any rate, easily summarized as call-return
pairs.
%
Note that jump compressed traces are not quite the same as
stack-traces as they show \emph{all} function calls, including
those that returned successfully.
%
For example, the display on the right in Figure~\ref{fig:factorial} shows
a jump-compressed trace, of size 4. The full trace for
the same example has size 19.

@@

% \item \ES{others?}
%
\begin{figure}[t]
\centering
% \includegraphics[width=\linewidth]{trace_size.pdf}
% \begin{minipage}{0.49\linewidth}
\includegraphics[width=\linewidth]{trace_size_step.pdf}
% \end{minipage}
% \begin{minipage}{0.49\linewidth}
\includegraphics[width=\linewidth]{trace_size_jump.pdf}
% \end{minipage}
\caption{Complexity of the generated traces. At least 90\% of the traces
  have a jump complexity of less than 10, with an average of 3.}
\label{fig:results-complexity}
\end{figure}
%

>>>>

\begin{figure}[t]
\centering
\begin{minipage}{\linewidth}
\includegraphics[width=\linewidth]{coverage.png}
\end{minipage}
\begin{minipage}{\linewidth}
\includegraphics[width=\linewidth]{distrib.png}
\end{minipage}
% \vspace{-8ex}
\caption{Results of our coverage testing and the distribution of test
  outcomes. Our random search successfully finds witnesses for 83\% of
  the programs in under one second, improving slightly to 87\% in under
  10 seconds. In both datasets we detect actual type errors about 82\%
  of the time, unbound variables or constructors 3-4\% of the time, and
  diverging loops 2-3\% of the time. For the remaining 11-12\% of the
  programs we are unable to provide any useful feedback.  }
\label{fig:results-witness}
\end{figure}

@@

\paragraph{Jump Compression}
%
A \emph{jump compressed} trace is one whose edges are limited to forward
or backward jumps. (The sizes of these traces is equal to the jump
metric below.)
%
In our experience, jump compression abstracts many details of the
computation that are often uninteresting or irrelevant to the
explanation, or at any rate, easily summarized as call-return
pairs.
%
Expanding the initial visualization to a jump-compressed trace is often
an ideal first step;
% expanding a jump-compressed traces ideal for the initial state of the
they provide a concise overview of the execution and allow the user to
step directly into an interesting function call, at which point the can
proceed with more fine-grained steps.
%
Note that jump compressed traces are not quite the same as
stack-traces as they show \emph{all} function calls, including
those that returned successfully.
%
For example, the display on the right in Figure~\ref{fig:factorial} shows
a jump-compressed trace, of size 4. The full trace for
the same example has size 19.

@@

% \item \ES{others?}
%
\begin{figure}[t]
\centering
\includegraphics[width=\linewidth]{trace_size_step.png}
\includegraphics[width=\linewidth]{trace_size_jump.png}
\caption{Complexity of the generated traces. At least 90\% of the traces
  have a jump complexity of less than 10, with an average of 3.}
\label{fig:results-complexity}
\end{figure}
%

### 2015-11-20 paper/pldi16-submission/introduction.tex
replace pdf figures with pngs

Type errors are a common stumbling block for students
trying to learn typed functional languages like \ocaml\
and \haskell.
%
Consider the ill-typed @fac@ function on the left in
Figure~\ref{fig:factorial}.
%
The function returns @true@ in the base case (instead of @1@),
and so \ocaml responds with the error message:
%
\begin{verbatim}
 File "fac.ml", line 3, characters 8-23:
 Error: This expression has type bool
        but an expression was expected
        of type int
\end{verbatim}
%
This message makes perfect sense to an expert who is familiar
with the language and has a good mental model of how the type
system works.
%
However, it may perplex a novice who has yet to develop such a
mental model.
%
To make matters worse, unification-based type inference algorithms
often report errors far removed from their source.
%
This further increases the novice's confusion and can actively mislead
them to focus their investigation on an irrelevant piece of code.
%
Much recent work has focused on analyzing unification constraints
to properly \emph{localize} a type error~\cite{lerner_searching_2007,chen_counter-factual_2014,zhang_toward_2014,pavlinovic_finding_2014},
but an accurate source location still does not explain \emph{why} the
program is wrong.

@@

In this paper we propose a new approach that explains
static type errors by \emph{dynamically} witnessing
how exactly an ill-typed program goes wrong.
%
We have developed \toolname, an interactive tool that uses
the source of the ill-typed function to automatically synthesize
the result on the right in Figure~\ref{fig:factorial}, which
shows how the recursive calls reduce to a configuration where
the program ``goes wrong'' --- \ie\ the @int@ value @1@ is to be
multiplied with the @bool@ value @true@.
We achieve this via three concrete contributions.

>>>>

Type errors are a common stumbling block for students
trying to learn typed functional languages like \ocaml\
and \haskell.
%
Consider the ill-typed @fac@ function on the left in
Figure~\ref{fig:factorial}.
%
The function returns @true@ in the base case (instead of @1@),
and so \ocaml responds with the error message:
%
\begin{verbatim}
 File "fac.ml", line 5, characters 8-17:
 Error: This expression has type bool
        but an expression was expected
        of type int
\end{verbatim}
%
This message makes perfect sense to an expert who is familiar
with the language and has a good mental model of how the type
system works.
%
However, it may perplex a novice who has yet to develop such a
mental model.
%
To make matters worse, unification-based type inference algorithms
often report errors far removed from their source.
%
This further increases the novice's confusion and can actively mislead
them to focus their investigation on an irrelevant piece of code.
%
Much recent work has focused on analyzing unification constraints
to properly \emph{localize} a type error~\cite{lerner_searching_2007,chen_counter-factual_2014,zhang_toward_2014,pavlinovic_finding_2014},
but an accurate source location still does not explain \emph{why} the
program is wrong.

@@

In this paper we propose a new approach that explains
static type errors by \emph{dynamically} witnessing
how an ill-typed program goes wrong.
%
We have developed \toolname, an interactive tool that uses
the source of the ill-typed function to automatically synthesize
the result on the right in Figure~\ref{fig:factorial}, which
shows how the recursive calls reduce to a configuration where
the program ``goes wrong'' --- \ie\ the @int@ value @1@ is to be
multiplied with the @bool@ value @true@.
We achieve this via three concrete contributions.
